\boldmath Primordial Black Hole Formation in a Dust Bouncing Model

### 2024-11-18 referee_response.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% My Packages %%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage{braket}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{physics,slashed}
\usepackage{graphicx}
\usepackage[toc,page]{appendix}
\usepackage{hyperref}
\usepackage[noabbrev]{cleveref}

\usepackage{xcolor}
\usepackage{mdframed}

\usepackage[hmarginratio=1:1]{geometry}

\begin{document}

We appreciate the referee’s careful review of our manuscript and the insightful feedback
provided. Below, we provide detailed responses to each point and explain the
corresponding changes made to the manuscript where applicable:

\begin{enumerate}
    \item
          \begin{enumerate}
              \item  To obtain the matter Lagrangian for this model, we use the Schutz
                    formalism, following Ref.~\cite{fluidgeral}. In this framework, the
                    Lagrangian corresponds to the fluid pressure, which depends on the
                    specific entropy and enthalpy. A detailed discussion of this
                    approach is provided in the cited reference. For clarity, we rewrote
                    the description of the Schutz formalism in Section 2.2 and ensured
                    that the references are correctly cited.

              \item Indeed, if the sound speed were to approach zero, we would encounter
                    a divergence problem. This is why we limit ourselves to small but
                    non-zero equations of state for dust, which is in agreement with the
                    literature. Regarding the $\zeta$ field, there are always field
                    configurations that would lead to non-perturbative behavior.
                    However, in the present context, given the initial conditions in
                    equation 3.32, we are considering the adiabatic vacuum for which
                    $\zeta \propto 1/\sqrt{z^2c_s} \propto \sqrt{c_s}$. Therefore, terms
                    of the form $z^2c_s^2\zeta^2$ are $\propto c_s$ while $\zeta^3
                        \propto \sqrt{c_s}^3$ This means that the perturbative expansion is valid
                    for small sound speeds, as discussed in
                    Ref.~\cite{vitenti2012large}.

                    We have reorganized Sections 4.3 and 4.4 and considered a very
                    small, but non-zero, equation of state for dark matter to analyze
                    the pressureless case.
          \end{enumerate}

    \item

          \begin{enumerate}
              \item We improved Sec.~2.2 by adding a more detailed explanation of the
                    Schutz formalism and the fluid Lagrangian. We also included a
                    discussion of the lapse function and how it can be defined a
                    posteriori for simplicity.

              \item In the improved Sec.~2.2, we have added a brief discussion of the
                    meaning of the effective energy density and pressure in the context
                    of the quantum corrections. In Ref.~\cite{fluidgeral}, the authors
                    show that the energy density and pressure appearing in the perturbed
                    equations are only the classical one computed with the Bohmian
                    trajectory. Therefore, the equation-of-state parameter is always
                    $w=p_d/\rho_d$.

              \item In the revised Sec.~2.2, we have included the effective equation
                    corresponding to the second Friedmann equation, now presented as Eq. 2.17:
                    $-2\dot{\bar{H}} = \bar{\rho}(1+w) - 2\rho_q.$
                    Notably, the right-hand side of this equation approaches zero around the
                    bounce. However, as discussed previously in Ref.~\cite{fluidgeral} it
                    has been shown that the energy density and pressure in the second-order
                    Lagrangian for the perturbations are only the classical components
                    evaluated along the Bohmian trajectory. Consequently, this term is never
                    zero, ensuring that the model is free from ghost instability issues.
              \item In Ref.~33 the author chooses a particular form for the wave
                    function at the bounce. This is indeed not a initial condition, but
                    a choice made to simplify the analysis. The same wave-function
                    propagated to $T\to-\infty$ would have an extra phase factor
                    proportional to the $q^2$, exactly like a the wave function of a
                    free Gaussian package in 1d. One can always start with a different
                    wave function and define the bounce as the time where the phase
                    factor proportional to $q^2$ is zero. Again, this is equivalent to a
                    free Gaussian package in 1d. Thus, in short, has to start at a large
                    but finite negative time, with a Gaussian wave-function containing a
                    phase factor proportional to $q^2$. Then, the bounce is defined as
                    the time where this phase factor is zero. We have included a brief
                    discussion of this point in the revised manuscript.
              \item In our approach, we use a perturbative method: the background is
                    solved first, and the Bohmian trajectory is derived from the
                    background wave function. Perturbations are then computed
                    conditioned on this Bohmian trajectory. In practice, during the
                    contraction phase, the significant part of the power spectrum is
                    generated in the classical regime, effectively following a
                    semi-classical approach. Quantum effects on the background become
                    relevant only near the bounce, where the relevant modes are on
                    super-Hubble scales. If the background is in a superposition of
                    Gaussian states, the perturbations would naturally reflect this
                    superposition. However, this aspect lies outside the scope of our
                    work, as we are focused here on black hole formation during the
                    contraction phase.

          \end{enumerate}

    \item In general, when both the background and the perturbations are quantized it is necessary
          to develop the second order Lagrangian for the perturbations without assuming the background
          is classical. Once this is done, one can show that for particular choices of the wave-function
          for the whole system (background plus perturbations) the perturbations can be treated as
          quantum fields evolving around the Bohmian trajectory. We included a brief discussion of this
          point in the revised manuscript, Sec.~3.3.

    \item In Ref.~\cite{vitenti2012large}, it is shown that for a Friedmann background with
          scalar perturbations, an appropriate gauge choice ensures the perturbative
          series remains valid in the bounce model. Nevertheless, this only shows that
          vacuum initial conditions around an initial Friedmann background are
          consistent with the perturbative expansion. The referee is correct in pointing
          out that the perturbative expansion may not be valid for more general initial
          conditions. In our study, we focus on this simpler model, leaving the
          exploration of a homogeneous and anisotropic background, specifically
          Bianchi-I models, for future work. A brief discussion of this possibility has
          been included in the conclusion.

    \item To simplify the presentation, we set the speed of light $c=1$ in the main
          text. We have now included a brief discussion of this choice in the revised
          manuscript. We also clarified the definition energy density and perturbation
          in the main text.
\end{enumerate}

### 2024-11-18 main.tex
Improving response and manuscript.

        In this section, we briefly discuss our adopted quantum bouncing background model, for which we follow mostly Refs.~\cite{nelson_peter_bouncing_original,nelson2021bouncing}. The primary motivation behind considering quantum bounces lies in the necessity to evade the primordial singularity at the end of the contracting phase. Classical General Relativity, as described by the Penrose-Hawking singularity theorems, requires the violation of the null energy condition $\rho + p \geq 0$ to avoid the singularity, a condition that often leads to instabilities in classical bouncing models.

@@

\subsection{Wheeler-DeWitt Equation Solutions}
    We start by applying the aforementioned considerations to a flat, homogeneous, and isotropic universe containing a single perfect fluid characterized by its pressure $\bar{p}$ and energy density $\bar{\rho}$, along with the equation of state $\bar{p} = w\bar{\rho}$, with a constant $w$. Subsequently, we specialize in the dust case where $w \approx 0$.

    \textcolor{red}{To proceed with the quantization, we first require the system's Hamiltonian, which can be derived from its Lagrangian. The Lagrangian is given by
    \begin{align}
        L &= L_g + L_m
    ,\end{align}
    where $L_g$ represents the gravitational component and $L_m$ denotes the matter contribution. These components are explicitly written as
    \begin{align}
    L_g &= \int d^3x \sqrt{-g}\frac{R}{2\kappa}
    \end{align}
    and
    \begin{align}
    \label{matterlag}
    L_m &= \int d^3x \sqrt{-g}\bar{p}
    .\end{align}
    In these expressions, $g$ is the determinant of the metric, $R$ is the curvature scalar, and $\kappa = \frac{8 \pi G}{c^4}$, with $c$ being the speed of light and $G$ the gravitational constant. The quantity $\bar{p}$ refers to the pressure of the fluid. For a complete derivation of the Lagrangian, we refer the reader to Ref.~\cite{fluidgeral}, while providing a brief outline of the key steps here.}

    \textcolor{red}{The gravitational Lagrangian is straightforwardly obtained by computing the Ricci scalar from the background metric, whose only degree of freedom is the scale factor. For the matter Lagrangian, we employ the Schutz formalism, as outlined in Ref.\cite{Schutz} and utilized in Ref.\cite{fluidgeral}. In this formalism, the specific enthalpy serves as the thermodynamic potential, defined as $v = (\bar{\rho} + \bar{\bar{p}})/n$, with $n$ being the number of particles.}

    \textcolor{red}{The fluid's enthalpy, in turn, depends on the pressure, entropy, and particle number. The first law of thermodynamics gives us the following relations:
    \begin{align}
        dv &= \tau ds + \frac{1}{n} d\bar{p}, \quad \rightarrow \quad \left.\frac{\partial \bar{p}}{\partial s}\right|_v = -\tau n, \quad \left.\frac{\partial \bar{p}}{\partial v}\right|_s = n
    ,\end{align}
    where $\tau$ is the temperature per particle and $s$ the specific entropy. This allows us to express the Lagrangian in Eq.~\eqref{matterlag}, with pressure as a function of enthalpy and entropy.}

    \textcolor{red}{To further determine the fluid's enthalpy, we decompose the velocity of the fluid into scalar potentials as follows:}
\textcolor{red}{
\begin{align} v_\mu &= \bar\nabla_\mu \varphi_1 + \varphi_2 \bar\nabla_\mu \varphi_3 + \varphi_4 \bar\nabla_\mu s ,\end{align}
where $\varphi_i$ (for $i = 1, 2, 3, 4$) are arbitrary scalar fields, $\bar{\nabla}_\mu$ is the background covariant derivative and $s \equiv \varphi_5$. The entalphy is then obtained as the modulus of this velocity. By varying the matter Lagrangian with respect to both the scalar fields and the metric tensor, we obtain the equations governing a perfect fluid and the conjugated momentum associated with the fluid's degree of freedom.}

In the above framework, the total Hamiltonian takes the form

\begin{align}
\label{total_hamiltonian_FLRW}
            H & = N\dpar{- \frac{\Pi_{\bar{a}}^{2} }{ 4\bar{a} } + \frac{\Pi_{T}}{\bar{a}^{3w}} } ,
        \end{align}
        where 
        \begin{align}
            \Pi_{\bar{a}} & \equiv \frac{ \partial L}{\partial \dot{\bar{a}}}
            ,\end{align}
        \begin{align}
            \Pi_T & \equiv \frac{ \partial L}{ \partial \dot{T}} 
        ,\end{align}
        are the canonical momenta associated with the scale factor $\bar{a}$ and $T$, 
        the latter being a variable that describes the fluid's evolution. Also, we denote the cosmic time derivative as $\frac{\partial q}{c\partial t} \equiv \dot{q}$, which has a dimension of one over length.
        \textcolor{red}{The momentum $\Pi_T$ is related to the fluid's degree of freedom and is given by $\Pi_T = c^{-1} \tilde{V} \bar{a}^{3(1 + w)} \bar{\rho}$, where $\tilde{V}$ represents the conformal spatial volume, and the fluid's density is a function of enthalpy and entropy.} Remarkably, $\Pi_T$ appears linearly in the Hamiltonian, and its classical equation of motion is given by
        $\dot{T} = { N }/{ \bar{a}^{3w} }$, implying $\dot{T} > 0$ and making $T$ a monotonic function of classical time. Hence, $T$ can be used as a time variable, allowing us to write a Schrödinger-like equation by choosing $N = \bar{a}^{3w} \implies \dot{T} = 1$. For the details, see Ref.~\cite{fluidgeral}.

        Canonical quantization is then performed by promoting classical variables to operators satisfying the canonical commutation relations. This process yields the following Wheeler-DeWitt equation for the wavefunction of the universe  $\Psi(\bar{a}, T)$ \cite{nelson_peter_bouncing_original}
                \begin{equation}
                \label{wdweq}
            i\frac{\partial}{\partial T}\Psi(q,T) = \frac{1}{4}\frac{\partial^{2}}{\partial q^{2}}\Psi(q,T),
        \end{equation}
       such that a specific operator factor ordering was chosen to preserve the symmetries of the classical system \cite{halliwell1990introductory} . Also we introduced the change of variable
        \begin{equation}
            q \equiv \frac{2\bar{a}^{\frac{3}{2}\dpar{1-w} } }{ 3\dpar{1-w } }
        .\end{equation}

Eq.~\eqref{wdweq} resembles a time-reversed free particle Schrödinger equation and, with appropriate boundary conditions, its solutions are wavefunctions of the scale factor $\bar{a}(T)$. We turn to the De Broglie-Bohm interpretation~\cite{mukhanov2005physical,nelson2000bohm}, such that assuming a Gaussian wavefunction $\Psi(\bar{a},T)$, the Bohmian trajectory solution reads
\begin{equation}
\label{bohm_scale_factor}
            \bar{a}(T) = \bar{a}_{B}\dcol{1 + \dpar{ \frac{T}{T_{ B}} }^2 }^{{\frac{1}{3}\frac{1}{\dpar{1-w}} } }  ,
        \end{equation}
        where $\bar{a}_{B}$ is an integration constant that represents the minimum scale factor value and $T_B$ is a small arbitrary constant related to the time scale of the bounce.  \textcolor{red}{Note that our model goes from $T=-\infty$ in the far past untill $T=0$ at the bounce such that we obtain a minimal scale factor at the bounce.} Remarkably, $\bar{a}(t) \neq 0$ for all $t$, which means that this model is non-singular and represents an eternal universe~\cite{nelson_peter_bouncing_original}. For a detailed derivation, see \cite{nelson2021bouncing}.

@@

        Eliminating the time $t$ using \eqref{bohm_time} in \eqref{bohm_hubble}, we find 
        \begin{equation}
        \label{friedman_scale}
            \bar{H}^{2} = \frac{ 4 }{ 9t^{2}_{b}}\dpar
{ \frac{ \bar{a}^{3}_{b} }{ \bar{a}^{3} } - \frac{\bar{a}_{b}^{6}}{ \bar{a}^{6} } }\, ,
        \end{equation}
        which is equivalent to the Friedmann equations
        \begin{equation}\label{effective_friedmann}
            \bar{H}^{2} = \frac{ \kappa c^{2} }{ 3 }\bar\rho - \bar{H}^{2}_{0}\Omega_{q0}\bar{a}^{-6},
        \end{equation}
        where $\bar\rho$ is the dust energy density. \textcolor{red}{Also, from the derivative of Eq.~\eqref{friedman_scale},
        \begin{align}
         c\dot{\bar{H}} = - \frac{\kappa c^{2}}{2} \bar\rho + 3\bar{H}^{2}_{0}\Omega_{q0}\bar{a}^{-6}    
        .\end{align}
        }
 In the above equations, there is an additional term $- \bar{H}^{2}_{0}\Omega_{q0}\bar{a}^{-6}$ when compared to the usual Friedmann equation for a classical universe. In our model, the scale factor dynamics mirrors a typical Friedmann equation with a total energy density $\rho_{T} = \bar\rho_{d} + \rho_{q}$, where $\rho_{d}$ represents the dust energy density and $\rho_{q} = -\Omega\rho_{c}\bar{a}^{-6}$ denotes an effective negative energy density accounting for quantum effects that circumvent the primordial singularity~\cite{vitenti2012large}. For large-scale factors $\bar{a} \gg \bar{a}_{b} $, Eq.~\eqref{effective_friedmann} and Eq.~\eqref{bohm_scale_factor} reduce to the classical dust dominated flat FLRW universe.
        \textcolor{red}{For instance, let us compute the time for which the quantum contribution is 10\% of the classical one. That means, from Eqs.~\eqref{friedman_scale} and \eqref{bohm_scale_factor} , that $\bar{a} = 0.1 a_b$ and $1 + \left(\frac{T}{T_B}\right)^2 = 1000$ for dust. In our dimensionless units, defined in Eq.~\eqref{tauparameter}, this would result in a time $\tau_a \approx -4.16$, which is extremely close to the bouncing point and it is not relevant for our analysis. Thus, effectively we still have a classical behavior with an energy density following the scale factor from the Bohm trajectory, and the quantum corrections only become relevant when $t \approx t_b$.}
         With this, we conclude the analysis of our background quantum bouncing model and move to its associated perturbations. In the next sections, we will consider the above quantum bouncing model with $w\approx 10^{-10}$ and $x_b = 10^{35}$ and we solve the equations numerically without any approximation.

@@

\subsection{Gauge Invariant Perturbations}
We follow mainly the perturbation theory developed in Refs.~\cite{covariant_bardeen, Vitenti2012}. Assuming only scalar perturbations, the total metric of the physical space-time is given by
\begin{align}
\label{physmetric}
    ds^2 = -(1- 2 \phi)c^2dt^2 + \bar{a}\bar{D}_i \mathcal{B} cdt dx^i + \bar{a}^2(1-2 \psi)\delta_{ij}dx^i dx^j - \bar{D}_i\bar{D}_j \mathcal{E} dx^i dx^j 
.\end{align}
Here, $\phi, \psi, \mathcal{B}$ and $\mathcal{E}$ denote the scalar metric perturbations that we will assume to be much smaller than one \cite{vitenti2012large}. Also, $\bar{D}_i$ is the spatial covariant derivative in the $i$-th direction. Note that our barotropic fluid has no anisotropic pressure.

@@

We define gauge invariant quantities by analyzing their transformations under gauge transformations, such that we can combine different gauge-dependent quantities to form new invariant ones~\cite{covariant_bardeen}. However, this leads to a freedom on the variable definitions since many combinations of variables may lead to gauge invariant quantities. With this in mind, we define the Bardeen gauge invariant variables~\cite{Bardeen1980}
\begin{align}
\label{bardeen1}
    \Phi &\equiv \phi + \dot{\delta \sigma}\, , \\
\label{bardeen2}
    \Psi &\equiv \psi - \frac{\Bar{H}\delta \sigma}{c}
,\end{align}
with
\begin{align}
    \delta\sigma = -(\dot{\mathcal{E}} - \mathcal{B}) + \frac{2 \Bar{H}\mathcal{E}}{c}
.\end{align}
It is important to notice that the new variables in Eqs.~\eqref{bardeen1}-\eqref{bardeen2}, and other gauge invariant variables do not have a physical meaning unless a gauge is chosen. For instance, in the case of the Newtonian gauge, $\delta\sigma =0$ and $\Phi$ represents the Newtonian potential.
Hence we must define our gauge invariant quantities such that they represent our desired physical variables when we assume a particular gauge choice.

We are mostly interested in the energy density perturbations that collapse to form PBHs. These perturbations can be examined using the density contrast, defined as
\begin{align}
	\label{densitycon}
	\delta &\equiv \frac{\delta \rho}{\bar{\rho} + \bar{p}}
,\end{align}
where $\delta\rho$ is the perturbation to the background energy density $\bar\rho$. The density contrast provides a normalized measurement of the energy density perturbation around the background matter density field. However, once again we are interested in its gauge invariant form
\begin{align}
\label{deltarhoinvariant}
    \tilde{\delta\rho} &\equiv \delta\rho - \mathcal{V} \dot{\Bar{\rho}}
,\end{align}
where $\mathcal{V}$ is the velocity perturbation of the fluid, which has its gauge invariant form
\begin{align}
    \tilde{\mathcal{V}} &\equiv \mathcal{V} + \delta\sigma
.\end{align}

>>>>

In this section, we briefly discuss our adopted quantum bouncing background model, for
which we follow mostly Refs.~\cite{nelson_peter_bouncing_original, fluidgeral,
	nelson2021bouncing}. The primary motivation behind considering quantum bounces lies in
the necessity to evade the primordial singularity at the end of the contracting phase.
Classical General Relativity, as described by the Penrose-Hawking singularity theorems,
requires the violation of the null energy condition $\rho + p \geq 0$ to avoid the
singularity, a condition that often leads to instabilities in classical bouncing models.

@@

\subsection{Wheeler-DeWitt Equation Solutions}

{\color{red} We start by applying the aforementioned considerations to a flat,
	homogeneous, and isotropic universe containing a single perfect fluid characterized by
	its pressure $\bar{p}$ and energy density $\bar{\rho}$, along with the equation of state
	$\bar{p} = w\bar{\rho}$, with a constant $w$. Subsequently, we specialize in the dust
	case where $w \approx 0$. In practice, we consider a barotropic fluid with a constant
	equation of state, following the Schutz formalism as described in
	Ref.~\cite{fluidgeral}. In this section we will briefly review the quantization of the
	flat-dust bouncing model, focusing on the background dynamics.

	To proceed with the quantization, we first derive the system's Hamiltonian
	from its corresponding Lagrangian. The Lagrangian for the system is given by
	\begin{align}\label{lagrangian}
		L & = \int \dd^3x \sqrt{-\bar{g}} \left[\frac{\bar{R}}{2\kappa}
			+ \bar{p}(\bar{v}, \bar{s})\right],
	\end{align}
	where, $\bar{g}$ denotes the determinant of the metric, $\bar{R}$ is the background
	Ricci scalar, and $\kappa = \frac{8 \pi G}{c^4}$, where $c$ is the speed of
	light and $G$ is the gravitational constant. We set $c = 1$ for simplicity. The
	term $\bar{p}$ represents the fluid pressure, which is expressed as a function of
	the specific enthalpy $\bar{v}$ and specific entropy $\bar{s}$.

	Using the Schutz formalism, the enthalpy can be decomposed into scalar potentials:
	\begin{align}
		\bar{v}_\mu & = \bar\nabla_\mu \varphi_1
		+ \varphi_2 \bar\nabla_\mu \varphi_3
		+ \varphi_4 \bar\nabla_\mu \bar{s},
		\quad \bar{v} \equiv \sqrt{-\bar{v}_\mu \bar{v}^\mu},
	\end{align}
	where $\varphi_i$ are four independent scalar potentials. Varying the Lagrangian with
	respect to these potentials and the specific entropy yields all thermodynamic relations,
	including the perfect fluid energy-momentum tensor, entropy conservation, and particle
	number conservation.

	Since we assume a barotropic fluid with a constant equation of state $\bar{p} = w
		\bar{\rho}$, it follows that $\partial \bar{p} / \partial \bar{s} \vert_{\bar{\rho}} =
		0$, where $\bar{\rho}$ is the energy density. This assumption simplifies the pressure to
	the form $\bar{p} = f(\bar{s}) \bar{v}^{(1+w)/w}$.

	After performing a Legendre transformation and solving the constraints, the Lagrangian
	simplifies to:
	\begin{align}
		L          & = \Pi_a \dot{\bar{a}} + \Pi_{\varphi_1} \dot{\varphi}_1 +
		N \left( \frac{\kappa \Pi_a^2}{12 V \bar{a}} - V \bar{a}^3 \bar{\rho} \right), \\
		\Pi_a      & = -\frac{6 V \bar{a} \dot{\bar{a}}}{N \kappa}, \qquad
		\Pi_{\varphi_1} = V \bar{a}^3 f \frac{1+w}{w}\bar{v}^{1/w},                    \\
		\bar{\rho} & = \frac{f}{w\bar{a}^{3(1+w)}} \left( \frac{w}{1+w}
		\frac{1}{Vf} \right)^{1+w}\Pi_{\varphi_1}^{1+w},
	\end{align}
	where $V$ is the conformal volume. Since the fluid term has the form
	$\Pi_{\varphi_1}^{1+w}$ and the momentum term for the scale factor includes a factor
	dependent on $\bar{a}$, we introduce a final canonical transformation to define new
	momenta, $\Pi_q$ and $\Pi_T$.
	\begin{align}
		\Pi_q & = \Pi_{\bar{a}} \bar{a}^{(-1+3w)/2},                                                 \\
		\Pi_T & = \frac{V f}{w} \left( \frac{w}{1+w} \frac{1}{Vf}\right)^{1+w} \Pi_{\varphi_1}^{1+w}
	\end{align}

	In the above framework, the total Hamiltonian takes the form
	\begin{align}
		\label{total_hamiltonian_FLRW}
		{\cal H} & = \frac{N}{\bar{a}^{3w}}\dpar{\Pi_{T}- \frac{\kappa\Pi_q^{2} }{12}} ,
	\end{align}
	Canonical quantization is then performed by promoting classical variables to operators
	satisfying the canonical commutation relations. This process yields the following
	Wheeler-DeWitt equation for the wave-function of the universe  $\Psi(\bar{a}, T)$
	(see Ref.~\cite{nelson_peter_bouncing_original}):
	\begin{equation}
		\label{wdweq}
		i\hbar\frac{\partial}{\partial T}\Psi(q,T) -
		\frac{\kappa\hbar^2}{12}\frac{\partial^{2}}{\partial q^{2}}\Psi(q,T)  = 0 ,
	\end{equation}
	such that a specific operator factor ordering was chosen to preserve the symmetries of
	the classical system \cite{halliwell1990introductory}. Also, note that our last
	canonical transformation lead to the following variable
	\begin{equation}
		q = \frac{2\bar{a}^{\frac{3}{2}\dpar{1-w} } }{ 3\dpar{1-w } }.
	\end{equation}
	In the WdW equation above, we circumvent the problem of time by interpreting the
	parameter $T$ as an intrinsic time variable that is monotonically related to the
	classical cosmic time in the classical limit. This allows us to interpret the equation
	as a Schrödinger-like equation.
}

Equation~\eqref{wdweq} resembles a time-reversed free particle Schrödinger equation and,
with appropriate boundary conditions, its solutions are wave-functions of the scale
factor $\bar{a}$. We turn to the De Broglie-Bohm
interpretation~\cite{mukhanov2005physical, nelson2000bohm}, such that assuming a Gaussian
wave-function $\Psi(q, T)$, the Bohmian trajectory solution translates to the scale
factor reads
\begin{equation}
	\label{bohm_scale_factor}
	\bar{a}(T) = \bar{a}_{B}\dcol{1 + \dpar{ \frac{T}{T_{ B}} }^2 }^{{\frac{1}{3}\frac{1}{\dpar{1-w}} } }  ,
\end{equation}
where $\bar{a}_{B}$ is an integration constant that represents the minimum scale factor
value and $T_B$ is a small arbitrary constant related to the time scale of the bounce.
	{\color{red} To simplify the analysis, we choose the lapse function $N = \bar{a}^{3w}$,
		but this choice is only possible after solving the Bohmian trajectory. The reason is
		that, initially, we cannot select $N$ as an operator function of the scale factor,
		However, once we have the Bohmian trajectory $\bar{a}(T)$, we can make this choice,
		using it purely for convenience in the analysis. In our model, $T$ extends from $T =
			-\infty$ in the far past to $T = 0$ at the bounce, where the scale factor $\bar{a}$
		reaches its minimum. } Remarkably, $\bar{a}(T) \neq 0$ for all $T$, which means that
this model is non-singular and represents an eternal
universe~\cite{nelson_peter_bouncing_original}. For a detailed derivation, see
\cite{nelson2021bouncing}.

{\color{red} It is important to clarify that in Ref.~\cite{nelson2021bouncing}, the
author selects a Gaussian wave function for the scale factor at the bounce. However,
this does not imply that this choice is the initial condition for our model. Generally,
a Gaussian wave packet of a one-dimensional free particle will develop a time-dependent
phase proportional to $q^2$. Thus, it is always possible to begin with a Gaussian wave
function at any time and define the bounce as the moment when this phase factor is zero.
For instance, in Ref.~\cite{Peter2016a}, the authors choose a Gaussian wave function at
any time and fix the phase using $H_0$. }

@@

Eliminating the time $t$ using \eqref{bohm_time} in \eqref{bohm_hubble}, we find
\begin{equation}
	\label{friedman_scale}
	\bar{H}^{2} = \frac{ 4 }{ 9t^{2}_{b}}\dpar
	{ \frac{ \bar{a}^{3}_{b} }{ \bar{a}^{3} } - \frac{\bar{a}_{b}^{6}}{ \bar{a}^{6} } }\, ,
\end{equation}
which is equivalent to the Friedmann equations
\begin{equation}\label{effective_friedmann}
	\bar{H}^{2} = \frac{\kappa}{ 3 }\bar\rho - \bar{H}^{2}_{0}\Omega_{q0}\bar{a}^{-6},
\end{equation}
where $\bar\rho$ is the dust energy density. {\color{red} Also, from the derivative of
		Eq.~\eqref{friedman_scale},
		\begin{align}
			\dot{\bar{H}} = - \frac{\kappa}{2} \bar\rho + 3\bar{H}^{2}_{0}\Omega_{q0}\bar{a}^{-6}.
		\end{align}
	}

In the above equations, there is an additional term
$-\bar{H}^{2}_{0}\Omega_{q0}\bar{a}^{-6}$ when compared to the usual Friedmann equation
for a classical universe. {\color{red} In our model, the scale factor dynamics resemble
		a typical Friedmann equation with a total energy density $\rho_{T} = \bar{\rho} +
			\rho_{q},$ where $\bar{\rho}$ is the dust energy density, and $\rho_{q} = -\Omega
			\rho_{c} \bar{a}^{-6}$ denotes an effective negative energy density with an equation of
		state $w_q = 1$ that accounts for quantum effects to avoid the primordial
		singularity~\cite{vitenti2012large}. It is important to note that $\rho_{q}$ is not a
		physical energy density but rather an effective one that emerges from the Bohmian. In
		the perturbative analysis, the energy density will always be associated only with
		$\bar{\rho}$. For large-scale factors $\bar{a} \gg \bar{a}_{b} $,
		Eq.~\eqref{effective_friedmann} and Eq.~\eqref{bohm_scale_factor} reduce to the
		classical dust dominated flat FLRW universe. With this, we conclude the analysis of our
		background quantum bouncing model and move to its associated perturbations. In the next
		sections, we will consider the above quantum bouncing model with $w\approx 10^{-10}$ and
		$\bar{a}(t_0)/a_b = 10^{35}$ where $t_0$ is the time during the contracting phase where
		$H(t_0) = -H_0$ for $H_0 = 70 \, \text{km/s/Mpc}$. Moreover, when solving the equations
		numerically, we use the exact expressions in terms of $w$ without approximation.}

@@

\subsection{Gauge Invariant Perturbations}

We follow mainly the perturbation theory developed in Refs.~\cite{covariant_bardeen,
	Vitenti2012}. Assuming only scalar perturbations, the total metric of the physical
space-time is given by
\begin{align}
	\label{physmetric-pert}
	\dd s^2 = -(1- 2 \phi)\dd t^2 + \bar{a}\bar{D}_i \mathcal{B} \dd t \dd x^i +
	\bar{a}^2(1-2 \psi)\delta_{ij}\dd x^i \dd x^j - \bar{D}_i\bar{D}_j \mathcal{E} \dd x^i \dd x^j
	.\end{align}
Here, $\phi, \psi, \mathcal{B}$ and $\mathcal{E}$ denote the scalar metric perturbations
that we will assume to be much smaller than one \cite{vitenti2012large}. Also,
$\bar{D}_i$ is the spatial covariant derivative in the $i$-th direction. Note that our
barotropic fluid has no anisotropic pressure.

	{\color{red} In the perturbative treatment, we continue to use the Schutz formalism
		and construct the Lagrangian as in Eq.~\eqref{lagrangian}, but now incorporating the
		perturbed metric. The perturbed Lagrangian and associated variables are defined as
		in Ref.~\cite{Vitenti2012}. Note that in the perturbed Lagrangian, the perturbations
		are always linked to the physical quantities and their corresponding background
		values. The effective energy density and pressure do not fluctuate; instead, the
		quantum effects from the background are incorporated through the use of the
		background Bohmian trajectories.}

@@

We define gauge invariant quantities by analyzing their transformations under gauge transformations, such that we can combine different gauge-dependent quantities to form new invariant ones~\cite{covariant_bardeen}. However, this leads to a freedom on the variable definitions since many combinations of variables may lead to gauge invariant quantities. With this in mind, we define the Bardeen gauge invariant variables~\cite{Bardeen1980}
\begin{align}
	\label{bardeen1}
	\Phi & \equiv \phi + \dot{\delta \sigma}\, , \\
	\label{bardeen2}
	\Psi & \equiv \psi - \Bar{H}\delta \sigma
	,\end{align}
with
\begin{align}
	\delta\sigma = -(\dot{\mathcal{E}} - \mathcal{B}) + {2 \Bar{H}\mathcal{E}}
	.\end{align}
It is important to notice that the new variables in
Eqs.~\eqref{bardeen1}-\eqref{bardeen2}, and other gauge invariant variables do not have
a physical meaning unless a gauge is chosen. For instance, in the case of the Newtonian
gauge, $\delta\sigma = 0$ and $\Phi$ represents the Newtonian potential. Hence we must
define our gauge invariant quantities such that they represent our desired physical
variables when we assume a particular gauge choice.

We are mostly interested in the energy density perturbations that collapse to form PBHs.
These perturbations can be examined using the density contrast, defined as
\begin{align}
	\label{densitycon}
	\delta & \equiv \frac{\delta \rho}{\bar{\rho} + \bar{p}}
	,\end{align}
where $\delta\rho$ is the perturbation to the background energy density $\bar\rho$. The
density contrast provides a normalized measurement of the energy density perturbation
around the background matter density field. However, once again we are interested in its
gauge invariant form
\begin{align}
	\label{deltarhoinvariant}
	\tilde{\delta\rho} & \equiv \delta\rho - \mathcal{V} \dot{\Bar{\rho}}
	,\end{align}
where $\mathcal{V}$ is the velocity perturbation of the fluid, which has its gauge
invariant form
\begin{align}
	\tilde{\mathcal{V}} & \equiv \mathcal{V} + \delta\sigma
	.\end{align}

Using the background Einstein equations, we can relate the density contrast to the
Bardeen variables and the gauge invariant curvature perturbation $\zeta$ through the
following relations~\cite{Mukhanov1992, Vitenti2013}
\begin{align}
	\Psi                                                    & =\Phi,
	\\
	\label{deltarho}
	-\frac{2\bar{D}^2 \Phi}{3  \kappa (\bar{\rho}+\bar{p})} & =   \frac{{\tilde{\delta\rho}}} {3(\bar{\rho}+\bar{p})}
	,
	\\
	\label{vrelation}
	\zeta                                                   & \equiv \Psi  + \bar{H} \tilde{\mathcal{V}}
	\\
	\label{zeta2}
	\zeta                                                   & = \frac{3  \bar{a}^3}{N^2z^2 c_s^2 } \left[\frac{\partial}{\partial t}\left(\frac{c\Phi }{3\bar{H}}\right) + \frac{\Phi}{3}\right]
	,                                                                                                                                                                                            \\
	\label{phiz}
	\bar{D}^{2} \Phi                                        & =  - \frac{z^2\bar{H}}{\bar{a}^3} \dot{\zeta} = -\frac{\bar{H}}{2\bar{a}^3} \Pi_\zeta
	.\end{align}
\textcolor{red}{In the above, $c^{2}_s=\dpar{ \frac{\partial\bar{\rho}}{\partial\bar{p}} }_{S}= w$ is the speed of sound, $\Pi_\zeta$ is the conjugated momenta related to $\zeta$ and
	\begin{align}
		\label{zdef}
		z^2=\frac{\kappa \bar{a}^3 (\bar{\rho} + \bar{p})}{2\bar{H}^2 c_{s}^2}
		.\end{align}}
For completeness, we also write the relation between the curvature perturbation and the
usual Mukhanov-Sasaki variable $v$, that is,
\begin{align}
	\label{msv}
	v & \equiv - \zeta z \sqrt{\frac{2}{\kappa}}
	.\end{align}
We will establish a connection between PBH formation and the density contrast in the next section. For now, it suffices to understand that the excess energy density associated with the perturbations leads to black hole formation, and we can measure such excess through the density contrast. Thus our definition in Eq.~\eqref{deltarhoinvariant} is extremely important and we will now analyze it.
